\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands
%\allsectionsfont{\normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{color}
\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines

\usepackage[compact]{titlesec}
\titlespacing{\section}{0pt}{*0}{*0}
\titlespacing{\subsection}{0pt}{*0}{*0}
\titlespacing{\subsubsection}{0pt}{*0}{*0}

\setlength{\headheight}{13.6pt} % Customize the height of the header
\setlength{\parskip}{\baselineskip}%
\setlength{\parsep}{0pt}
\setlength{\headsep}{0pt}
\setlength{\topskip}{0pt}
\setlength{\topmargin}{0pt}
\setlength{\topsep}{0pt}
\setlength{\partopsep}{0pt}

%\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
%\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
%\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Department of Computer Science, Rochester Institute of Technology} \\ [25pt] % Your university, school and/or department name(s)
\horrule{2pt} \\[0.4cm] % Thin top horizontal rule
\huge Performance Evaluation of $k-SAT$ Solvers\\ 
\Large Applied to Graph Arrowing \\
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Christopher Wood \\ Advisor: Professor Stanis{\l}aw Radziszowski} % Your name

\date{\normalsize \today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

\section{Introduction}
In his 1972 seminal paper entitled, ``Reducibility Among Combinatorial Problems,'' Karp introduced a 
list of 21 NP-complete problems, including Boolean satisfiability, the maximum cut of a graph, 
and 0-1 integer programming \cite{karp72}. The complexity of these problems was proven by deriving
a polynomial-time reduction from CIRCUIT-SAT = $\{\langle C \rangle : C$ is a satisfiable Boolean combinational
circuit $\}$, the first problem shown to be NP-complete by Cook in 1971 \cite{cook71-np}, starting
the rush of complexity theory research. 

The problem $3-SAT$, or more formally, $3-CNF SAT$, is a special case of satisfiability.
It is a decision problem in which takes as input a 3-CNF Boolean formula and returns YES if 
the formula is satisfiable, and NO otherwise \cite{clrs90-algorithms}. A 3-CNF formula, more formally known as a Boolean 
formula in 3-conjunctive normal form, is expressed as the Boolean AND of arbitrarily many clauses,
where each clause is the Boolean OR of three literals, which is a Boolean variable or its negation. 
Such a Boolean formula is said to be satisfiable if and only if there exists an assignment of truth 
values to the variables such that substituting them into the literals of the formula will cause it 
to evaluate to true (or 1). Expressed as a formal language, we have that 
$3-SAT = \{\langle \phi \rangle : \phi \text{ is satisfiable }\}$.

In 2002 Hans van Maaren of and John Franco initiated the public SAT competition in
search of optimal performing SAT solvers judged by a variety of criteria and specializations, 
including their ability to demonstrate satisfiability and exhaustively prove unsatisfiability.
In addition, since $SAT$ is a problem that often arises in academia and the industry, 
each of the candidate solvers are rigorously tested with massive application-specific, 
crafted, and random Boolean formulas as input. With three different solver
specializations tested against three different types of inputs, and a first, second, 
and third place awarded to the candidates, a total of $27$ possible trophies are
awarded each year. In the most recent competition held in 2011, solvers were
tested using CPU time and world-clock time as a basis for their results, thus expanding
the trophy space to $54$ slots. The next competition is slated to take place in 2013. 

%TODO: sat problem, sat competition, categories of evaluation, types of problems, graph arrowing...
The $SAT$ problem is particularly interesting when applied to graph arrowing. It can be shown
how to reduce the question $G \overbrace{\to}^\text{?} \to (3,3)^e$ to an equivalent $3-CNF$ formula
$\phi_G$ such that $G \not \to (3,3)^e \Leftrightarrow \phi_G$. Intuitively, this is a very promising
technique for determining if $G \to (3,3)^e$ for $K_4$-free graphs $G$.

The immediate application of this technique is to attack the upper bound of the Folkman number
$F_e(3,3;4)$. In particular, to lower this bound, we will need to find a 
graph $G$ on $n$ vertices where $G \to (3,3)^e$. It has been conjectured that 
$G_{127} = G(127,3) = (\mathbb{Z}_{127}, E = \{(x,y) | x - y = \alpha^3 \mod 127\})$ is a
prime candidate for witnessing an upper bound of $127$ because of its denseness and large number
of triangles. To determine whether $G_{127}$ is indeed a witness we will decompose 
the problem of arrowing into problems on subgraphs $H$ that witness $H \not \to (3,3)^e$, 
and then carefully extend $H$ to encompass all of $G$. For each subgraph $H$ will generate
a corresponding $3$-$CNF$ formula $\phi_G$ by mapping the edges in $E(H)$ to variables 
in $\phi_H \in$ 3-SAT, and for edge adding the following clauses to $\phi_H$:
\begin{align*}
(x + y + z) \wedge (\bar{x} + \bar{y} + \bar{z})
\end{align*}
If $H$ can be extended to encompass all of $G$ and $\phi_G$ is 
shown to be unsatisfiable, then $G \to (3,3)^e$, and so $F_e(3,3;4) = 127$.

Unfortunately, while this approach seems simple upfront, the complexity of the formulas
$\phi_H$ with $n \approx 85$ has proven to be very difficult for modern SAT solvers
to handle. In this study, we will attempt to determine the structure of these formulas
that makes them so difficult to solve. We will also present a comprehensive performance 
comparison for the popular SAT solvers entered into the SAT competition, including
Minisat, zChaff, glucose, ppfolio //, ppfolio seq, contrasat hack, and 3S.

\section{$k-SAT$ Algorithms}
{\color{red} TODO}: most solvers use a Minisat variation - articulate that here for completeness

\section{Selected $k-SAT$ Solvers}
{\color{red} TODO}: discuss each 
%TODO: minisat, zchaff, glucose, ppfolio //, ppfolio seq, contrasat hack, 3S

\section{Performance}
{\color{red} TODO}: determine test cases first
%TODO: performance results will go here

%%%%% REFERENCES %%%%%

\begin{thebibliography}{9}

\bibitem{karp72} Richard Karp. Reducibility among combinatorial problems. \emph{Complexity of Computer Computations, (RE Miller and JM Thatcher, eds.)} (1972), 85–103.

\bibitem{cook71-np} Stephen A. Cook. The Complexity of Theorem-Proving Procedures. \emph{In Proceedings of the third annual ACM symposium on Theory of computing (STOC '71)}. ACM, New York, NY, USA (1971), 151-158. DOI=10.1145/800157.805047. {\tt http://doi.acm.org/10.1145/800157.805047}

\bibitem{clrs90-algorithms} Gormen, Thomas H., Charles E. Leiserson, Ronald L. Rivest, and Clifford Stein. Introduction to Algorithms. MIT Press 44 (1990), 97-138.

\bibitem{minisat} Niklas S\"{o}rensson and Niklas E\`{e}n. Minisat v1.13 - A SAT Solver with Conflict-Clause Minimization. \emph{SAT 2005} (2005) 53.

\bibitem{zchaff} Yogesh Mahajan, Zhaohui Fu, and Sharad Malik. Zchaff2004: An Efficient SAT Solver. \emph{Theory and Applications of Satisfiability Testing}. Springer Berlin/Heidelberg (2005).

\end{thebibliography}

\end{document}